Automated Assessment of Residual Plots with Computer Vision Models

### 2025-06-16 paper.tex
Adjust Examples, Limitation and Future Works and Conclusions sections

In this section, we present the performance of the trained computer
vision model on three example datasets, including the dataset associated
with the residual plot displaying a ``left-triangle'' shape (Figure
\ref{fig:false-finding}), the Boston housing dataset
\citep{harrison1978hedonic}, and the ``dino'' datasets from the from the
\texttt{datasauRus} package \citep{datasaurus}.

@@

In Section \ref{sec-model-introduction}, we presented an example
residual plot showcased in Figure \ref{fig:false-finding}, illustrating
how humans might misinterpret the ``left-triangle'' shape as indicative
of heteroskedasticity. Additionally, the Breusch-Pagan test yielded a
rejection with a \(p\)-value of 0.046, despite the residuals originating
from a correctly specified model. Figure \ref{fig:false-lineup} offers a
lineup for this fitted model, showcasing various degrees of
``left-triangle'' shape across all residual plots. This phenomenon is
evidently caused by the skewed distribution of the fitted values.
Notably, if the residual plot were evaluated through a visual test, it
would not be rejected since the true residual plot positioned at 10 can
not be distinguished from the others.

Figure \ref{fig:false-check} presents the results of the assessment by
the computer vision model. Notably, the observed visual signal strength
is considerably lower than the 95\% sample quantile of the null
distribution. Moreover, the bootstrapped distribution suggests that it
is highly improbable for the fitted model to be misspecified as the
majority of bootstrapped fitted models will not be rejected. Thus, for
this particular fitted model, both the visual test and the computer
vision model will not reject \(H_0\). However, the Breusch-Pagan test
will reject \(H_0\) because it can not effectively utilize information
from null plots.

The attention map at Figure \ref{fig:false-check}B suggests that the
estimation is highly influenced by the top-right and bottom-right part
of the residual plot, as it forms two vertices of the triangular shape.
A principal component analysis (PCA) is also performed on the output of
the global pooling layer of the computer vision model. As mentioned in
\citet{simonyan2014very}, a computer vision model built upon the
convolutional blocks can be viewed as a feature extractor. For the
\(32 \times 32\) model, there are 256 features outputted from the global
pooling layer, which can be further used for different visual tasks not
limited to distance prediction. To see if these features can be
effectively used for distinguishing null plots and true residual plot,
we linearly project them into the first and second principal components
space as shown in Figure \ref{fig:false-check}D. It can be observed that
because the bootstrapped plots are mostly similar to the null plots, the
points drawn in different colors are mixed together. The true residual
plot is also covered by both the cluster of null plots and cluster of
bootstrapped plots. This accurately reflects our understanding of Figure
\ref{fig:false-lineup}.

@@

{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/false-check-1}

@@

The Boston housing dataset, originally published by
\citet{harrison1978hedonic}, offers insights into housing in the Boston,
Massachusetts area. For illustration purposes, we utilize a reduced
version from Kaggle, comprising 489 rows and 4 columns: average number
of rooms per dwelling (RM), percentage of lower status of the population
(LSTAT), pupil-teacher ratio by town (PTRATIO), and median value of
owner-occupied homes in \$1000's (MEDV). In our analysis, MEDV will
serve as the response variable, while the other columns will function as
predictors in a linear regression model. Our primary focus is to detect
non-linearity, because the relationships between RM and MEDV or LSTAT
and MEDV are non-linear.

Figure \ref{fig:boston-check} displays the residual plot and the
assessment conducted by the computer vision model. A clear non-linearity
pattern resembling a ``U'' shape is shown in the plot A. Furthermore,
the RESET test yields a very small \(p\)-value. The estimated distance
\(\hat{D}\) significantly exceeds \(Q_{null}(0.95)\), leading to
rejection of \(H_0\). The bootstrapped distribution also suggests that
almost all the bootstrapped fitted models will be rejected, indicating
that the fitted model is unlikely to be correctly specified. The
attention map in plot B suggests the center of the image has higher
leverage than other areas, and it is the turning point of the ``U''
shape. The PCA provided in plot D shows two distinct clusters of data
points, further underling the visual differences between bootstrapped
plots and null plots. This coincides the findings from Figure
\ref{fig:boston-lineup}, where the true residual plot exhibiting a ``U''
shape is visually distinctive from null plots. If a visual test is
conducted by human, \(H_0\) will also be rejected.

@@

{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/boston-check-1}

@@

The computer vision model possesses the capability to detect not only
typical issues like non-linearity, heteroskedasticity, and non-normality
but also artifact visual patterns resembling real-world objects, as long
as they do not appear in null plots. These visual patterns can be
challenging to categorize in terms of model violations. Therefore, we
will employ the RESET test, the Breusch-Pagan test, and the Shapiro-Wilk
test \citep{shapiro1965analysis} for comparison.

The ``dino'' dataset within the \texttt{datasauRus} R package
exemplifies this scenario. With only two columns, x and y, fitting a
regression model to this data yields a residual plot resembling a
``dinosaur'', as displayed in Figure \ref{fig:dino-check}A.
Unsurprisingly, this distinct residual plot stands out in a lineup, as
shown in Figure \ref{fig:dino-lineup}. A visual test conducted by humans
would undoubtedly reject \(H_0\).

According to the residual plot assessment by the computer vision model,
\(\hat{D}\) exceeds \(Q_{null}(0.95)\), warranting a rejection of
\(H_0\). Additionally, most of the bootstrapped fitted models will be
rejected, indicating an misspecified model. However, both the RESET test
and the Breusch-Pagan test yield \(p\)-values greater than 0.3, leading
to a non-rejection of \(H_0\). Only the Shapiro-Wilk test rejects the
normality assumption with a small \(p\)-value.

More importantly, the attention map in Figure \ref{fig:dino-check}B
clearly exhibits a ``dinosaur'' shape, strongly suggesting that the
distance prediction is based on human-perceptible visual patterns. The
computer vision model effectively captures the contour or outline of the
embedded shape, similar to how humans interpret residual plots.
Additionally, the PCA in Figure \ref{fig:dino-check}D demonstrates that
the cluster of bootstrapped plots is positioned at the corner of the
cluster of null plots.

In practice, without accessing the residual plot, it would be
challenging to identify the artificial pattern of the residuals.
Moreover, conducting a normality test for a fitted regression model is
not always standard practice among analysts. Even when performed,
violating the normality assumption is sometimes deemed acceptable,
especially considering the application of quasi-maximum likelihood
estimation in linear regression. This example underscores the importance
of evaluating residual plots and highlights how the proposed computer
vision model can facilitate this process.

@@

{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/dino-check-1}

@@

Despite the computer vision model performing well with general cases
under the synthetic data generation scheme and the three examples used
in this paper, this study has several limitations that could guide
future work.

The proposed distance measure assumes that the true model is a classical
normal linear regression model, which can be restrictive. Although this
paper does not address the relaxation of this assumption, there are
potential methods to evaluate other types of regression models. The most
comprehensive approach would be to define a distance measure for each
different class of regression model and then train the computer vision
model following the methodology described in this paper. To accelerate
training, one could use the convolutional blocks of our trained model as
a feature extractor and perform transfer learning on top of it, as these
blocks effectively capture shapes in residual plots. Another approach
would be to transform the residuals so they are roughly normally
distributed and have constant variance. If only raw residuals are used,
the distance-based statistical testing compares the difference in
distance to a classical normal linear regression model for the true
residual plot and null plots. This comparison is meaningful only if the
difference can be identified by the distance measure proposed in this
paper.

There are other types of residual plots commonly used in diagnostics,
such as residuals vs.~predictor and quantile-quantile plots. In this
study, we focused on the most commonly used residual plot as a starting
point for exploring the new field of automated visual inference.
Similarly, we did not explore other, more sophisticated computer vision
model architectures and specifications for the same reason. While the
performance of the computer vision model is acceptable, there is still
room for improvement to achieve behavior more closely resembling that of
humans interpreting residual plots. This may require external survey
data or human subject experiment data to understand the fundamental
differences between our implementation and human evaluation.

@@

In this paper, we have introduced a distance measure based on
Kullback-Leibler divergence to quantify the disparity between the
residual distribution of a fitted classical normal linear regression
model and the reference residual distribution assumed under correct
model specification. This distance measure effectively captures the
magnitude of model violations in misspecified models. We propose a
computer vision model to estimate this distance, utilizing the residual
plot of the fitted model as input.

Moreover, the estimated distance enables the development of a formal
statistical testing procedure by evaluating a large number of null plots
generated from the fitted model. Additionally, employing bootstrapping
techniques and refitting the regression model allows us to ascertain how
frequently the fitted model is considered misspecified if data were
repeatedly obtained from the same data generating process.

The trained computer vision model demonstrates strong performance on
both the training and test sets, although it exhibits slightly lower
performance on residual plots with non-linearity visual patterns
compared to other types of violations. The statistical tests relying on
the estimated distance predicted by the computer vision model exhibit
lower sensitivity compared to conventional tests but higher sensitivity
compared to visual tests conducted by humans. While the estimated
distance generally mirrors the strength of the visual signal perceived
by humans, there remains scope for further improvement in its
performance.

Several examples are provided to showcase the effectiveness of the
proposed method across different scenarios, emphasizing the similarity
between visual tests and distance-based tests. Overall, both visual
tests and distance-based tests can be viewed as ensemble of tests,
aiming to assess any violations of model assumptions collectively. In
contrast, individual residual diagnostic tests such as the RESET test
and the Breusch-Pagan test only evaluate specific violations of model
assumptions. In practice, selecting an appropriate set of statistical
tests for regression diagnostics can be challenging, particularly given
the necessity of adjusting the significance level for each test.

Our method holds significant value as it helps alleviate a portion of
analysts' workload associated with assessing residual plots. While we
recommend analysts to continue reading residual plots whenever feasible,
as they offer invaluable insights, our approach serves as a valuable
tool for automating the diagnostic process or for supplementary purposes
when needed.

>>>>

In this section, we present the performance of the trained computer
vision model on three example datasets, including the dataset associated
with the residual plot displaying a ``left-triangle'' shape, the Boston
housing dataset \citep{harrison1978hedonic}, and the ``dino'' datasets
from the from the \texttt{datasauRus} package \citep{datasaurus}.

@@

In Section \ref{sec-model-introduction}, we introduced a residual plot
(Figure \ref{fig:false-finding}) where the ``left-triangle'' shape could
be misinterpreted as evidence of heteroskedasticity. Although the
residuals are drawn from a correctly specified model, the Breusch-Pagan
test rejects \(H_0\) (\(p = 0.046\)). This visual artifact stems from
the skewed distribution of fitted values, as illustrated in the lineup
in Figure \ref{fig:false-lineup}, where similar ``left-triangle''
patterns appear across all residual plots. Since the true plot (at
position 10) is visually indistinguishable from the null plots, a visual
test would not reject \(H_0\).

The computer vision model yields a consistent conclusion. As shown in
Figure \ref{fig:false-check}, the estimated distance \(\hat{D}\) falls
well below the 95\% quantile of the null distribution, with a
\(p\)-value of \(0.33\). The bootstrapped distribution further suggests
that rejection of the fitted model is highly unlikely. Together, the
visual and model-based tests do not reject \(H_0\), in contrast to the
Breusch-Pagan test, which lacks the context provided by null plots and
may be misled by visual artifacts.

To further interpret the model's output, the attention map in Figure
\ref{fig:false-check}B, computed as the gradient of the output with
respect to the greyscale input, shows that the top-right and
bottom-right corners of the residual plot have the most influence,
corresponding to the triangle's vertices. In addition, a principal
component analysis (PCA) of the 256 features extracted by the model's
global pooling layer (Figure \ref{fig:false-check}D) shows that null and
bootstrapped plots cluster together in the space defined by the first
two principal components, and the true residual plot is covered by the
cluster. This overlap reinforces the conclusion that the true plot is
visually indistinct, aligning with our interpretation of the lineup in
Figure \ref{fig:false-lineup}.

@@

{\centering \includegraphics[width=0.8\linewidth]{paper_files/figure-latex/false-check-1}

@@

The Boston housing dataset, originally published by
\citet{harrison1978hedonic}, provides insights into housing in the
Boston, Massachusetts area. For illustration, we use a reduced Kaggle
version containing 489 observations and four variables: average number
of rooms per dwelling (RM), percentage of lower status population
(LSTAT), pupil-teacher ratio by town (PTRATIO), and median home value
(MEDV). We fit a linear regression model with MEDV as the response and
the remaining variables as predictors, focusing on detecting
non-linearity, because the relationships between RM and MEDV or LSTAT
and MEDV are non-linear.

Figure \ref{fig:boston-check} summarizes the model evaluation. The
residual plot (Figure \ref{fig:boston-check}A) reveals a distinct
``U''-shaped pattern, suggesting non-linearity. The RESET test strongly
rejects \(H_0\) with a very small \(p\)-value, and the computer vision
model yields a large estimated distance \(\hat{D}\), well above the 95\%
quantile of the null distribution (\(p\text{-value} = 0.00498\)). The
bootstrapped distribution also shows widespread rejection, indicating
the model is likely misspecified. The attention map (Figure
\ref{fig:boston-check}B) highlights the central region of the residual
plot, corresponding to the turning point of the ``U'', as most
influential. In the PCA projection (Figure \ref{fig:boston-check}D),
bootstrapped and null plots form two distinct clusters, emphasizing the
visual separability. This aligns with the lineup in Figure
\ref{fig:boston-lineup}, where the true plot stands out clearly. A human
visual test would also likely reject \(H_0\) in this case.

@@

{\centering \includegraphics[width=0.8\linewidth]{paper_files/figure-latex/boston-check-1}

@@

Beyond detecting common issues like non-linearity, heteroskedasticity,
and non-normality, the computer vision model can also identify unusual
visual artifacts, such as shapes resembling real-world objects, as long
as such patterns do not appear in null plots. These cases are
challenging to categorize using conventional model diagnostics, so we
compare the model's output with results from the RESET test,
Breusch-Pagan test, and Shapiro-Wilk test \citep{shapiro1965analysis}.

The ``dino'' dataset from the \texttt{datasauRus} R package illustrates
this situation. It contains only two variables, \(x\) and \(y\), but
fitting a regression model produces a residual plot that clearly
resembles a dinosaur (Figure \ref{fig:dino-check}A). This pattern is
visually distinctive in the lineup shown in Figure
\ref{fig:dino-lineup}, where a human visual test would undoubtedly
reject \(H_0\).

Consistent with this, the computer vision model assigns a distance
\(\hat{D}\) above the 95\% quantile of the null distribution
(\(p-\text{value} = 0.00498\)), leading to rejection of \(H_0\). The
bootstrapped distribution further supports this, with most fitted models
also being rejected. In contrast, the RESET and Breusch-Pagan tests
yield \(p\)-values above \(0.3\), suggesting no violation. Only the
Shapiro-Wilk test rejects normality with a small \(p\)-value.

Crucially, the attention map in Figure \ref{fig:dino-check}B highlights
the dinosaur shape, indicating that the model's decision is driven by
human-perceptible features. The model appears to capture contours and
outlines in a manner similar to human visual interpretation. In the PCA
plot (Figure \ref{fig:dino-check}D), the cluster of bootstrapped plots
is positioned at the corner of the cluster of null plots, isolated from
the cluster of null plots.

In practice, such visual artifacts would be difficult to detect without
inspecting the residual plot directly. Furthermore, analysts do not
always test for normality, and even when violations are found, they are
often deemed acceptable due to the robustness of linear regression under
quasi-likelihood theory. This example highlights the value of visual
diagnostics and the unique role of the proposed computer vision model in
supporting them.

@@

{\centering \includegraphics[width=0.8\linewidth]{paper_files/figure-latex/dino-check-1}

@@

While the computer vision model performs well under the synthetic data
generation scheme and across the three illustrative examples, this study
has several limitations that point to directions for future work.

First, the proposed distance measure assumes the true model follows a
classical normal linear regression framework, which may be restrictive.
Although we do not explore relaxation of this assumption here, future
work could extend the approach to other types of regression models. One
possibility is to define a distance measure tailored to each model class
and re-train the computer vision model accordingly. To reduce
computational cost, the convolutional blocks from our current model
could be reused via transfer learning, as it is already effective at
capturing visual features. Alternatively, transforming residuals to
approximate normality and constant variance could allow for meaningful
comparisons under our current distance measure. However, if raw
residuals are used, the interpretation remains valid only when the
visual differences between the true and null plots are detectable by the
proposed distance measure.

Second, this study focused solely on the standard residuals vs.~fitted
values plot, leaving out other common diagnostic plots such as residuals
vs.~predictors and Q-Q plots. These alternative plots may offer
additional information and could be incorporated into a broader visual
testing framework in future research.

Finally, we chose a relatively simple computer vision architecture to
establish a baseline for automated visual inference. While the current
model shows promise, there is room to improve its alignment with human
judgment. Achieving this may require external data from surveys or
controlled experiments to better understand how human observers
interpret residual plots and how these judgments differ from the model's
outputs.

@@

In this paper, we have introduced a distance measure based on the
Kullback-Leibler divergence to quantify disparity between the residual
distribution of a fitted classical normal linear regression model and
the reference distribution under correct specification. This measure
captures the extent of model violations and forms the basis of a
computer vision model that estimates the distance using residual plots
as input.

The estimated distance supports a formal statistical testing procedure
by comparing the true residual plot against null plots generated from
the fitted model. Through bootstrapping and model refitting, we also
evaluate how often a correctly specified model would be flagged as
misspecified under repeated sampling.

The trained computer vision model performs well on both training and
test data, though its performance is somewhat lower for non-linear
patterns. Statistical tests based on the model's predicted distance show
lower sensitivity than conventional tests but higher sensitivity than
human visual tests. Importantly, the predicted distance broadly aligns
with the strength of visual signals perceived by humans, though further
improvements are possible.

We demonstrate the method across several scenarios, highlighting its
effectiveness and its conceptual alignment with visual inference. Both
visual and distance-based tests act as omnibus procedures, detecting a
wide range of model violations collectively. In contrast, conventional
tests such as RESET and Breusch-Pagan target specific assumptions,
making them less flexible and potentially harder to combine without
adjusting for multiple testing.

Overall, our approach offers a practical tool to support regression
diagnostics by reducing the manual effort required for residual plot
inspection. While we recommend analysts to continue reading residual
plots whenever feasible, our method can augment the process, either as
an automated screening tool or a supplement to human judgment.
